---
title: Notes on non-standard evaluation in R
description: 
author: Victor Yuan
date: 2026-02-30
draft: true
categories:
  - R
---

Notes on Hadley's Advanced R book.

# Chapter 17 

https://adv-r.hadley.nz/meta-big-picture.html

## `expr` vs `enexpr`

`expr` captures code as it is written

`enexpr` captures code that a user has passed to a function


```{r}
library(rlang)
capture_it <- function(x) {
  expr(x)
}
capture_it(1 + 2 + 3)
```



```{r}
capture_it <- function(x) {
  enexpr(x)
}
capture_it(1 + 2 + 3)
```

Terminology: `enexpr` allows "quoting" user supplied expressions to functions. The quoted user expression can itself be then modified.

## Constructing expressions programmatically

Expressions can be constructed with `rlang::call2`


```{r}
call2('+', 1, 2)
```


`!!` operator takes the abstract syntax tree (AST) and injects it into a function


```{r}
square_it <- function(x) {
  captured <- enexpr(x)
  expr(!!captured * !!captured)
}
square_it(1 + 2 + 3)
```


```{r}
xx <- expr(x + x)
yy <- expr(y + y)

expr(!!xx / !!yy)
#> (x + x)/(y + y)
```

## Evaluating code

In above examples, we "captured" or "quoted" code. To evaluate code, means to reaplce the variables and functions using a namespace/environment, and then execute the resulting AST.


```{r}
eval(expr(x + y), env(x = 1, y = 2))
```

What if no env is provided?

```{r}

eval(expr(x + y))
```

This is how tidyverse functions work, they capture user supplied code and then evaluate in an environment where objects/variables are the columns of a data frame.


```{r}
string_math <- function(x) {
  e <- env(
    caller_env(),
    `+` = function(x, y) paste0(x, y),
    `*` = function(x, y) strrep(x, y)
  )

  eval(enexpr(x), e)
}

name <- "Hadley"
string_math("Hello " + name)
#> [1] "Hello Hadley"
string_math(("x" * 2 + "-y") * 3)
#> [1] "xx-yxx-yxx-y"
```

## enquo vs enexpr

`enexpr` captures code user supplies to function

`enquo` captures code user supplies to function + the environment 

`enquo` is used with `eval_tidy` in the context of a data mask ie in a dataframe operation


# 18

## 18.2

### 18.2.1

#> █─f 
#> └─█─g 
#>   └─█─h

f(g(h()))

#> █─`+` 
#> ├─█─`+` 
#> │ ├─1 
#> │ └─2 
#> └─3
(1+2)+3

#> █─`*` 
#> ├─█─`(` 
#> │ └─█─`+` 
#> │   ├─x 
#> │   └─y 
#> └─z

z*(x+y)
